---
title: "Soil preservation scores mapped to SGDBE"
format: html
---

Here we follow Kibblewhite et al 2015 and redo their steps to connect their soil preservation scores derived from soil type to the STUs of the SGDBE.

They write: "SGDBE consists of (1) a geometric dataset at scale 1:1,000,000 and (2) a semantic dataset containing attribute files. The geometrical component of the database are polygons that form soil mapping units (SMUs) (EC, 2003). SMUs contain one or more STUs that form a discrete landscape unit with shared characteristics. The semantic dataset of the SGDBE describe the soil types present within the SMU and the percentage of the SMU area occupied by each STU present within it. Thus multiple attribute values are linked to the polygons of the geometrical dataset. *We assigned the appropriate preservation numeric score to each STU in each SMU and calculated the spatially weighted mean of these scores for each SMU*. Areas of artificial surfaces, permanent ice or snow cover and water bodies were excluded from the analysis. We then mapped the mean scores for the SMUs across the European Union after assigning colours to a continuous scale of mean preservation score. The percentage areas of soil across the European Union that provide good (highest third of scale), fair and poor (lowest third of scale) preservation capacities for bones, teeth and shells; organic materials; and metals (Cu, bronze and Fe) were then calculated. "

-\> therefore we will 
1. get the STUs downloaded from SGDBE 
2. connect STU to soil preservation scores depending on soil types 
3. save file 
4. add STU to SMU via percentage of soil type 
5. save file 
6. take administrative boundaries dataset and cut SMUs to different levels of admin unit 
7. calculate percentage of SMU per area of administrative unit 
8. for LAU level: only keep the largest value 
9. for NUTS levels: keep percentages of preservation scores per admin unit

<!-- for russia and EURASIA see: https://docs.isric.org/soil-geographic-databases/sections/soil-geographic-databases-asia-eurasia.html#eurasia -->

## 1. already published data (shape file by Kibblewhite et al)

already published for the EU under: Kibblewhite M., Tóth G. and Hermann T. 2015 . "Maps of preservation capacity of cultural artefacts and buried materials in soils in the EU", dataset/maps downloaded from the European Soil Data Centre, esdac.jrc.ec.europa.eu.

Kibblewhite et al 2015, 252:
Numeric values were assigned to each preservation category (poor = 1, fair =
2, good = 3). These values were assigned by relating the STU descrip-
tion (e.g., soil profile, diagnostic horizons and soil properties) to soil
properties that drive material destruction (e.g., hydrology and associat-
ed redox conditions, pH, base saturation, soil organic matter content,
chloride levels)

```{r}
#| label: load data and packages

library(ggplot2)
library(ggspatial)
library(viridis)
library(sf)
library(dplyr)


kw_scores <- read_sf(dsn = "./geodata/SF_cultural/SF_cultural.shp") # percentage, soil ID and preservation scores

load("./geodata/ne_map_ESTER.RData")

ne_map <- read_sf(dsn = "./geodata/ne_10m_admin_1_states_provinces.shp") # world background map
```


```{r}
#| label: prepare for mapping


## set all to the same CRS

# kw_scores_4326 <- st_transform(kw_scores, crs = 4326)

kw_scores_3035 <- st_transform(kw_scores, crs = 3035) ## probably better to use projected data

ne_map_3035 <- st_transform(ne_map, crs = 3035)

kw_scores$SOIL_ID[!st_is_valid(kw_scores)] # several units don't seem to be valid

kw_scores_3035 <- kw_scores_3035 |> filter(SOIL_ID > 0)

ne_map_ESTER_3035 <- st_transform(ne_map_ESTER, crs = 3035) 

bbox <- st_bbox(ne_map_ESTER_3035)
bbox <- st_as_sfc(bbox)

sf_use_s2(FALSE) # needs to be run once, there's a bug, see https://github.com/r-spatial/sf/issues/1762 

kw_scores_3035 <- st_make_valid(kw_scores_3035) # needs to run after sf_use_s2

kw_scores_ESTER <- st_intersection(kw_scores_3035, ne_map_ESTER_3035)

```


### Bone preservation
```{r}

# poor = 1, fair = 2, good = 3

## scale to our system from 0 (no bias = good) to -3
### first scale to 0 to 3 and add minus after

# formula is y = new min + (x - old min)(new max - new min) / (old max - old min)
# new max is 0 , new min is -3, old min is 1, old max is 3

#y = -3 + (x - 1)(0 - (-3) / (3 - 1)

# -3 + (x-1)*3/2

kw_scores_ESTER$bones[kw_scores_ESTER$bones == 0] <- NA

kw_scores_ESTER$score_bones_new <- (-3 + (kw_scores_ESTER$bones - 1) * 3/2)

```



```{r}


ggplot()+
  geom_sf(data = ne_map_3035, colour = "grey", fill = "white", size = 0.5, alpha = 0.5) + # world background
  geom_sf(data = kw_scores_ESTER, aes(fill = score_bones_new, colour = score_bones_new))+
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         pad_x = unit(0.2, "in"), pad_y = unit(0.2, "in"),
                         style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = st_coordinates(bbox)[c(1,2),1], # min & max of x values
           ylim = st_coordinates(bbox)[c(2,3),2]) + # min & max of y values
  labs(fill ="",
       colour ="",
       title = "Bias for preservation of bone, shell etc.",
       caption = "weighted mean, based on Kibblewhite et al. 2015")+
  scale_fill_viridis_c(na.value = NA)+
  scale_colour_viridis_c(na.value = NA)+
  theme(legend.key = element_rect(fill = "white",
                                  colour = "white",
                                  linewidth = 0, linetype = "solid"),
        panel.background = element_rect(fill = "lightblue",
                                        colour = "lightblue",
                                        linewidth = 0.5, linetype = "solid")) 



ggsave("./maps/preservation_bones.png", dpi = 300, width = 20, height = 20, units = "cm")
```

### Metals



```{r}

kw_scores_ESTER$metals[kw_scores_ESTER$metals == 0] <- NA

kw_scores_ESTER$score_metals_new <- (-3 + (kw_scores_ESTER$metals - 1) * 3/2)


ggplot()+
  geom_sf(data = ne_map_3035, colour = "grey", fill = "white", size = 0.5, alpha = 0.5) + # world background
  geom_sf(data = kw_scores_ESTER, aes(fill = score_metals_new, color = score_metals_new))+
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         pad_x = unit(0.2, "in"), pad_y = unit(0.2, "in"),
                         style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = st_coordinates(bbox)[c(1,2),1], # min & max of x values
           ylim = st_coordinates(bbox)[c(2,3),2]) + # min & max of y values
  labs(fill ="",
       colour ="",
       title = "Bias for preservation of metals",
       caption = "weighted mean, based on Kibblewhite et al. 2015")+
  scale_fill_viridis_c(na.value = NA)+
  scale_colour_viridis_c(na.value = NA)+
  theme(legend.key = element_rect(fill = "white",
                                  colour = "white",
                                  linewidth = 0, linetype = "solid"),
        panel.background = element_rect(fill = "lightblue",
                                        colour = "lightblue",
                                        linewidth = 0.5, linetype = "solid")) 

ggsave("./maps/preservation_metals.png", dpi = 300, width = 20, height = 20, units = "cm")
```

### Organics



```{r}

kw_scores_ESTER$organics[kw_scores_ESTER$organics == 0] <- NA

kw_scores_ESTER$score_organics_new <- (-3 + (kw_scores_ESTER$organics - 1) * 3/2)

ggplot()+
 geom_sf(data = ne_map_3035, colour = "grey", fill = "white", size = 0.5, alpha = 0.5) + # world background
  geom_sf(data = kw_scores_ESTER, aes(fill = score_organics_new, color = score_organics_new))+
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         pad_x = unit(0.2, "in"), pad_y = unit(0.2, "in"),
                         style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = st_coordinates(bbox)[c(1,2),1], # min & max of x values
           ylim = st_coordinates(bbox)[c(2,3),2]) + # min & max of y values
  labs(fill ="",
       colour ="",
       title = "Bias for preservation of organics",
       caption = "weighted mean, based on Kibblewhite et al. 2015")+
  scale_fill_viridis_c(na.value = NA)+
  scale_colour_viridis_c(na.value = NA)+
  theme(legend.key = element_rect(fill = "white",
                                  colour = "white",
                                  linewidth = 0, linetype = "solid"),
        panel.background = element_rect(fill = "lightblue",
                                        colour = "lightblue",
                                        linewidth = 0.5, linetype = "solid")) 

ggsave("./maps/preservation_organics.png", dpi = 300, width = 20, height = 20, units = "cm")
```

### Stratigraphy preservation


```{r}

kw_scores_ESTER$strati[kw_scores_ESTER$strati == 0] <- NA

kw_scores_ESTER$score_strati_new <- (-3 + (kw_scores_ESTER$strati - 1) * 3/2)


ggplot()+
  geom_sf(data = ne_map_3035, colour = "grey", fill = "white", size = 0.5, alpha = 0.5) + # world background
  geom_sf(data = kw_scores_ESTER, aes(fill = score_strati_new, color = score_strati_new))+
  annotation_scale(location = "bl", width_hint = 0.5) +
#  annotation_north_arrow(location = "tl", which_north = "true", 
 #                        pad_x = unit(0.2, "in"), pad_y = unit(0.2, "in"),
  #                       style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = st_coordinates(bbox)[c(1,2),1], # min & max of x values
           ylim = st_coordinates(bbox)[c(2,3),2]) + # min & max of y values
  labs(fill ="",
       colour ="",
       title = "Bias for preservation of stratigraphy",
       caption = "weighted mean, based on Kibblewhite et al. 2015")+
  scale_fill_viridis_c(na.value = NA)+
  scale_colour_viridis_c(na.value = NA)+
  theme(legend.key = element_rect(fill = "white",
                                  colour = "white",
                                  linewidth = 0, linetype = "solid"),
        panel.background = element_rect(fill = "lightblue",
                                        colour = "lightblue",
                                        linewidth = 0.5, linetype = "solid")) 

ggsave("./maps/preservation_stratigraphy.png", dpi = 300, width = 20, height = 20, units = "cm")
```


... BUT: 
- missing Switzerland, Russia and other non-EU-areas, that have been mapped in the SGDBE


## 2. SGDBE data for Russia: connect STU to soil preservation scores depending on soil types

```{r}
#| label: load data
#install.packages("foreign") to read dbf files

library(foreign)
dis <- read.dbf("/media/sophie/Data/ESTER_Nextcloud/Daten/EuropeanSoilDatabase/soilDB_shapefiles_and_attributes/stu_sgdbe.dbf", as.is = F)

stuorg <- read.dbf("/media/sophie/Data/ESTER_Nextcloud/Daten/EuropeanSoilDatabase/soilDB_shapefiles_and_attributes/stuorg.dbf", as.is = F)

smu <- read.dbf("/media/sophie/Data/ESTER_Nextcloud/Daten/EuropeanSoilDatabase/soilDB_shapefiles_and_attributes/smu_sgdbe.dbf", as.is = F)
```


```{r}
#| label: get values for WRBFU classes
#| 
wrbfu_values <- unique(dis$WRBFU) # sollte die WRBFU - Klassifizierung sein

library(stringr)

wrbfu_NAvalues <- wrbfu_values[str_detect(wrbfu_values, pattern = "^[0-9]+")]
#wrbfu_NAvalues <- as.character(wrbfu_NAvalues)


wrbfu_values_class <- as.data.frame(wrbfu_values)

wrbfu_values_class <- wrbfu_values_class |>
  filter(!wrbfu_values %in% wrbfu_NAvalues)

wrbfu_values_class2 <- tidyr::extract(wrbfu_values_class, wrbfu_values, c("ref_group", "sub_group"), "([A-Z\\s]+)([a-z]+)")
```


```{r}
#| label: extract European soil classifications from PDF

# The European Soil Databases (ESDB), distribution version v2.0 is freely available to the public after user registration. Please note that the spatial data are in vector format.
##  year 2004 -> documentation 

## get information from documentation


library(tabulapdf)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(janitor)
library(stringr)


pdf_file <- "./pdf_data/WRB_2007_deutsche_Ausgabe_Böden.pdf" # load PDF of German WRB 2006

# get_page_dims(pdf_file) # dimensions to get an idea of page size ->  595 842

## page 141

# region as a vector containing the top, left, bottom and right coordinates

## reference groups
region <- c(170, 0, 250, 595)

tab <- extract_tables(pdf_file, 
               pages = 141,
               guess = FALSE,
               col_names = F,
               method = "stream",
               area = list(region) 
               )[[1]]


ref_groups <- data.frame(data.table::rbindlist(list(tab[,1:2],tab[,3:4], tab[,5:6], tab[,7:8])))



## qualifier
region <- c(260, 0, 690, 595)

tab2 <- extract_tables(pdf_file, 
               pages = 141,
               guess = FALSE,
               col_names = F,
               method = "stream",
               area = list(region) 
               )[[1]]


qualifier <- data.frame(data.table::rbindlist(list(tab2[,1:2],tab2[,3:4], tab2[,5:6], tab2[,7:8])))

qualifier <- qualifier |>
  filter(!is.na(X1))

all_groups <- cross_join(ref_groups, qualifier) # this is hypothetical, I know

#all_groups$X1.x[all_groups$X1.x == "Leptsol"] <- "Leptosol"

all_groups <- all_groups |>
  mutate(name = paste(X1.y, X1.x),
         short = paste0(X2.x, X2.y)) 

all_groups$match <- tolower(all_groups$name)
```

```{r}
#| label: match with Kibblewhite system

load("./data/appendix_kibblewhite_etal_2015_A+B.RData")

appendix$match <- tolower(appendix$`Name in WRB 1998`)
## WRB 1998 seems to be a better match for soils in SGDBE

match <- left_join(appendix, all_groups)

# a few records don't match, but all in all this worked fine
```

## 3. save

```{r}
save(match, file = "./data/soil_WRB_classification_preservation_scores_1998.RData")
```

```{r}
#| label: data cleaning

## fill in missing values in "short" by hand and use 1998 classification -> seems to be more differentiated and results in more matches

match_filled <- match |>
  mutate(short = case_when(
    is.na(short) & match == "anthrosol" ~ "AN",
    is.na(short) & `Name in WRB 1998` == "Histosol" ~ "HS",
    is.na(short) & `Name in WRB 1998` == "Chernozem" ~ "CH",
    is.na(short)  &`Name in WRB 1998` == "Haplic Leptsol"  ~ "LPha",
    TRUE ~ short) )

## in two cases (ARha and ARpr) climate and wetness can differ -> merged to be only one row

match_filled <- match_filled |>
  mutate(Wetness = case_when(
    short == "ARha" ~ "Dry/Moist",
    short == "ARpr" ~ "Dry/Moist",
    TRUE ~ Wetness),
    TypicalClimate = case_when(
      short == "ARha"~ "Dry/Moist",
      short == "ARpr" ~ "Dry/Moist",
      TRUE ~ TypicalClimate  )) |>
  unique()

save(match_filled, file = "./data/soil_WRB_classification_preservation_scores.RData")

```


## map the STUs to SMU 

4.  add Kibblewhite infos to STU 

```{r}
#| label: load data STU and SMU and kibblewhite scores matched with WRB classes
 
load("./data/soil_WRB_classification_preservation_scores.RData")

library(sf)

stuorg <- read_sf(dsn = "./geodata/EuropeanSoilDatabase/soilDB_shapefiles_and_attributes/stuorg.dbf") # STU ID to SMU ID with percentage of area

stu_sgdbe <- read_sf(dsn = "./geodata/EuropeanSoilDatabase/soilDB_shapefiles_and_attributes/stu_sgdbe.dbf") # STU with all infos (also WRBFU)
country <- read_sf(dsn = "./geodata/EuropeanSoilDatabase/soilDB_shapefiles_and_attributes/country.shp")

sgdbe4_0 <- read_sf(dsn = "./geodata/EuropeanSoilDatabase/soilDB_shapefiles_and_attributes/sgdbe4_0.shp") # DOM STU and its percentage per SMU

stu_kw <- left_join(stu_sgdbe, match_filled, by = c("WRBFU" = "short"))

```

```{r}
#| label: check for missing information

list_WRBFU_stu <- unique(stu_sgdbe$WRBFU)
list_WRBFU_kw <- unique(match_filled$short)

missing_wrb_in_kw <- list_WRBFU_stu[!(list_WRBFU_stu %in% list_WRBFU_kw)] # 38 missing soil types in Kibblewhite's dataset :-( 

missing_wrb_in_stu <- list_WRBFU_stu[!(list_WRBFU_kw %in% list_WRBFU_stu)] # only 8, also not as important
```

There are 38 soil types in the WRBFU classification of the SGDBE that don't match with any Kibblewhite group. 

Following the assumption that if all subtypes of a reference group share one preservation score in the Kibblewhite et al dataset, other subtypes would follow the same pattern, their scores are assigned to the subtypes missing in the Kibblewhite et al data.


```{r}
#| label: match the missing classes with other scores of ref group

## "ANth"     "CMhu"     "CRca"     "LPum"     "ABeun"    "CRum"  
## "CMge"     "CHgs"     "CHch"     "GLcc"     "GLso"     "CRgl"     "CRhi"     "FLum"     "HScy"     "HSsz"     "PZhi"     "CRcc"    
## "CRha"     "CRtu"     "CRoa"     "CRye"     "ANao"     "ANhi"     "CRan"     "LPmo"     "PLlv"     "PLgl"     "PLmo"     "PLab"    
## "CLha"     "CLszn"    "RGge"     "VRgl"     "SC"      

soil_groups <- match_filled |> select(X1.x, X2.x, X1.y, X2.y, name) |> unique()
colnames(soil_groups) <- c("ref_gr", "ref_group_short", "sub_gr", "sub_group_short", "name")


library(tidyr)

stu_kw$WRBFU[stu_kw$WRBFU %in% c("1 1 1" ,"2 2 22 2", "3 3 33 3", "4 4 44 4" ,"5 5 55 5" ,"6 6 66 6")]  <- NA

stu_kw <- stu_kw |>
  separate_wider_position(cols = WRBFU, c(ref_group_short = 2, sub_group_short = 2 ), 
                          too_few = "align_start",
                          too_many = "debug",
                          cols_remove = F)
  
stu_kw <- left_join(stu_kw, soil_groups)

# check for bones: 

ref_group_bones <- table(match_filled$X2.x, match_filled$`Bones etc.`)

stu_kw <- stu_kw |>
  mutate(`Bones etc.` = case_when(
    is.na(`Bones etc.`) & ref_group_short == "AB" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "AC" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "AL" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "AN" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "AR" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "AT" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "CL" ~ "Good",
    is.na(`Bones etc.`) & ref_group_short == "CH" ~ "Fair",
    is.na(`Bones etc.`) & ref_group_short == "GY" ~ "Good",
    is.na(`Bones etc.`) & ref_group_short == "KS" ~ "Fair",
    is.na(`Bones etc.`) & ref_group_short == "PH" ~ "Fair",
    is.na(`Bones etc.`) & ref_group_short == "PZ" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "SC" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "SN" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "UM" ~ "Poor",
    is.na(`Bones etc.`) & ref_group_short == "VR" ~ "Poor",
    TRUE ~ `Bones etc.`
  ))

stu_kw_b_na <- stu_kw |> filter(is.na(`Bones etc.`)) |> select(WRBFU) |> unique()
stu_kw_b_na$WRBFU

# check for metals: 

ref_group_metals <- table(match_filled$X2.x, match_filled$Metals)

stu_kw <- stu_kw |>
  mutate(Metals = case_when(
    is.na(Metals) & ref_group_short == "AB" ~ "Poor",
    is.na(Metals) & ref_group_short == "AC" ~ "Poor",
    is.na(Metals) & ref_group_short == "AL" ~ "Poor",
    is.na(Metals) & ref_group_short == "AN" ~ "Poor",
    is.na(Metals) & ref_group_short == "AR" ~ "Poor",
    is.na(Metals) & ref_group_short == "AT" ~ "Poor",
    is.na(Metals) & ref_group_short == "CL" ~ "Good",
    is.na(Metals) & ref_group_short == "FL" ~ "Poor",
    is.na(Metals) & ref_group_short == "CH" ~ "Fair",
    is.na(Metals) & ref_group_short == "GY" ~ "Good",
    is.na(Metals) & ref_group_short == "KS" ~ "Fair",
    is.na(Metals) & ref_group_short == "PZ" ~ "Poor",
    is.na(Metals) & ref_group_short == "RG" ~ "Poor",
    is.na(Metals) & ref_group_short == "SC" ~ "Poor",
    is.na(Metals) & ref_group_short == "SN" ~ "Poor",
    is.na(Metals) & ref_group_short == "UM" ~ "Poor",
    is.na(Metals) & ref_group_short == "VR" ~ "Poor",
    TRUE ~ Metals
  ))

stu_kw_m_na <- stu_kw |> filter(is.na(Metals)) |> select(WRBFU) |> unique()
stu_kw_m_na$WRBFU


# check for organics: 

ref_group_organics <- table(match_filled$X2.x, match_filled$Organics)

stu_kw <- stu_kw |>
  mutate(Organics = case_when(
    is.na(Organics) & ref_group_short == "AB" ~ "Fair",
    is.na(Organics) & ref_group_short == "AC" ~ "Poor",
    is.na(Organics) & ref_group_short == "AL" ~ "Poor",
    is.na(Organics) & ref_group_short == "AN" ~ "Poor",
    is.na(Organics) & ref_group_short == "AR" ~ "Fair",
    is.na(Organics) & ref_group_short == "AT" ~ "Fair",
    is.na(Organics) & ref_group_short == "CH" ~ "Fair",
    is.na(Organics) & ref_group_short == "CL" ~ "Good",
    is.na(Organics) & ref_group_short == "GY" ~ "Good",
    is.na(Organics) & ref_group_short == "HS" ~ "Good",
    is.na(Organics) & ref_group_short == "KS" ~ "Fair",
    is.na(Organics) & ref_group_short == "PH" ~ "Fair",
    is.na(Organics) & ref_group_short == "PL" ~ "Fair",
    is.na(Organics) & ref_group_short == "SC" ~ "Poor",
    is.na(Organics) & ref_group_short == "SN" ~ "Poor",
    is.na(Organics) & ref_group_short == "VR" ~ "Poor",
    TRUE ~ Organics
  ))

stu_kw_o_na <- stu_kw |> filter(is.na(Organics)) |> select(WRBFU) |> unique()
stu_kw_o_na$WRBFU

# check for stratigraphy: 

ref_group_stratigraphy <- table(match_filled$X2.x, match_filled$Stratigraphy)

stu_kw <- stu_kw |>
  mutate(Stratigraphy = case_when(
    is.na(Stratigraphy) & ref_group_short == "AB" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "AC" ~ "Poor",
    is.na(Stratigraphy) & ref_group_short == "AL" ~ "Poor",
    is.na(Stratigraphy) & ref_group_short == "AN" ~ "Poor",
    is.na(Stratigraphy) & ref_group_short == "AR" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "AT" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "CH" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "CL" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "FL" ~ "Good",
    is.na(Stratigraphy) & ref_group_short == "GL" ~ "Good",
    is.na(Stratigraphy) & ref_group_short == "GY" ~ "Good",
    is.na(Stratigraphy) & ref_group_short == "HS" ~ "Good",
    is.na(Stratigraphy) & ref_group_short == "KS" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "LP" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "PH" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "PL" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "PZ" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "RG" ~ "Poor",
    is.na(Stratigraphy) & ref_group_short == "SC" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "SN" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "UM" ~ "Fair",
    is.na(Stratigraphy) & ref_group_short == "VR" ~ "Poor",
    TRUE ~ Stratigraphy
  ))

# check for NAs

stu_kw_s_na <- stu_kw |> filter(is.na(Stratigraphy)) |> select(WRBFU) |> unique()
stu_kw_s_na$WRBFU
```

Completely missing is the group of Cryosols (CR), which are soils influenced by ice / frost (as described by Kibblewhite et al, they were removed from the dataset).

In case of organics, the following reference groups where **not** used to inform missing values, as there were differences in the subgroups (`unique(stu_kw_o_na$WRBFU)`): CM, LP, GL, FL, PZ, RG

In case of metals, the following reference groups where **not** used to inform missing values, as there were differences in the subgroups (`unique(stu_kw_m_na$WRBFU)`): CM, LP, GL, HS, LP, PL

In case of bones, the following reference groups where **not** used to inform missing values, as there were differences in the subgroups (`unique(stu_kw_b_na$WRBFU)`): CM, LP, GL, FL, HS, LP, PL, RG

In case of stratigraphy, the following reference groups where **not** used to inform missing values, as there were differences in the subgroups (`unique(stu_kw_s_na$WRBFU)`): CM

```{r}

## scores by Kibblewhite et al so far only character, needs to be factor:

stu_kw$`Bones etc.` <- factor(stu_kw$`Bones etc.`, ordered = T,
                              levels = c("Good", "Fair", "Poor"))


stu_kw$Organics <- factor(stu_kw$Organics, ordered = T,
                              levels = c("Good", "Fair", "Poor"))

stu_kw$Metals <- factor(stu_kw$Metals, ordered = T,
                              levels = c("Good", "Fair", "Poor"))

stu_kw$Stratigraphy <- factor(stu_kw$Stratigraphy, ordered = T,
                              levels = c("Good", "Fair", "Poor"))

save(stu_kw, file = "./data/STU_preservation_scores.RData")

```

5. to SMU via percentage of soil type (weighted mean) ... or connect SMU to match via soil type?

```{r}

stu_smu <- left_join(stuorg, stu_kw, by = "STU")


stu_smu_b <- stu_smu |>
  group_by(SMU) |>
  mutate(score_bones =  sum(PCAREA * as.numeric(`Bones etc.`) / 100),
         score_metals =  sum(PCAREA * as.numeric(Metals) / 100),
         score_organics = sum(PCAREA * as.numeric(Organics) / 100),
         score_strati = sum(PCAREA * as.numeric(Stratigraphy) / 100))|>
  select(SMU, contains("score")) |>
  unique()

smu_map <- left_join(sgdbe4_0, stu_smu_b)

#smu_map$score_bones <- as.numeric(smu_map$score_bones ) #nötig??
#smu_map$score_metals <- as.numeric(smu_map$score_metals ) 
#smu_map$score_organics <- as.numeric(smu_map$score_organics ) 
#smu_map$score_strati <- as.numeric(smu_map$score_strati ) 

## scale to our system from 0 (no bias = good) to -3
### first scale to 0 to 3 and add minus after

# formula is y = new min + (x - old min)(new max - new min) / (old max - old min)


## other version:
# new min is -3 , new max is 0, old min is 1, old max is 3
## ergo: y = -3 + (x-1)*(0- (-3))/(3-1) = -3 + (x-1)*3/2

smu_map$score_bones_new <- -3 + ((smu_map$score_bones-1)*3/2)
smu_map$score_metals_new <- -3 + ((smu_map$score_metals-1)*3/2)
smu_map$score_organics_new <- -3 + ((smu_map$score_organics-1)*3/2)
smu_map$score_strati_new <- -3 +((smu_map$score_strati-1)*3/2) 

```

```{r}
library(ggplot2)
library(ggspatial)

ggplot()+
  geom_sf(data = ne_map_3035, colour = "grey", fill = "white", size = 0.5, alpha = 0.5) + # world background
  geom_sf(data = smu_map, aes(fill = score_bones_new, colour = score_bones_new))+
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         pad_x = unit(0.2, "in"), pad_y = unit(0.2, "in"),
                         style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = st_coordinates(bbox)[c(1,2),1], # min & max of x values
           ylim = st_coordinates(bbox)[c(2,3),2]) + # min & max of y values
  labs(fill ="",
       colour ="",
       title = "Bias for preservation of bone, shell etc.",
       caption = "weighted mean, based on Kibblewhite et al. 2015")+
  scale_fill_viridis_c(na.value = NA)+
  scale_colour_viridis_c(na.value = NA)+
  theme(legend.key = element_rect(fill = "white",
                                  colour = "white",
                                  linewidth = 0, linetype = "solid"),
        panel.background = element_rect(fill = "lightblue",
                                        colour = "lightblue",
                                        linewidth = 0.5, linetype = "solid")) 

ggsave("./maps/preservation_bones_own_calc.png",  dpi = 300, width = 20, height = 20, units = "cm")
```

## check differences to maps provided by Kibblewhite

```{r}
#| label: join both datasets
# join the dataframes

## 1. cut smu map to bbox of ESTER and transform dataset to INSPIRE system: https://epsg.io/3035
#smu_map_3035 <- st_transform(smu_map, crs = 3035)

### for checking with old scores
smu_map_3035 <- left_join(sgdbe4_0, stu_smu_b)
smu_map_3035 <- st_transform(smu_map_3035, crs = 3035)

sf_use_s2(FALSE) # needs to be run once, there's a bug, see https://github.com/r-spatial/sf/issues/1762 

smu_map_3035 <- st_make_valid(smu_map_3035) # needs to run after sf_use_s2

smu_map_3035 <- smu_map_3035 |> filter(SOIL_ID > 0)

smu_map_ESTER <- st_intersection(smu_map_3035, ne_map_ESTER_3035)

comparison <- st_join(smu_map_ESTER, kw_scores_ESTER)

comparison |>
ggplot()+
    geom_point(aes(x = bones, y = score_bones ))


```


```{r}
## cut a few columns to make it easier to handle

smu_map_ESTER_b <- smu_map_ESTER |> select(SOIL_ID, contains("score"))

kw_scores_ESTER_b <- kw_scores_ESTER |> select(SOIL_ID, bones, metals, organics, strati, iso_a2, iso_3166_2, iso_3166_c, contains("score"))

kw_scores_ESTER_b <- st_transform(kw_scores_ESTER_b, crs = 3035) # vmtl nicht mehr nötig?

comparison <- st_join(smu_map_ESTER_b, kw_scores_ESTER_b)

comparison$bones_comp <- comparison$score_bones_new.x - comparison$score_bones_new.y
summary(comparison$bones_comp)



```


...

5.  save file
6.  take administrative boundaries dataset and cut SMUs to different levels of admin unit
7.  calculate percentage of SMU per area of administrative unit
8.  for LAU level: only keep the largest value
9.  for NUTS levels: keep percentages of preservation scores per admin unit
